Gravitational Lens Recovery with \Glass: How to measure the mass profile and shape of a lens

### 2013-06-13 limits/ms.tex
Put usenatbib in \documentclass to avoid variable redefinition problem on Jonathan's machine. Fix duplicate label.
further hacking on the introduciton. Looking more like it now but still a bit clunky ...

With this coming explosion in lens data, it is timely to re-asses the impact of model degeneracies on quantities of interest like the circularly averaged density profile of a lens or its shape. Some significant work has already proceeded in this direction in the literature. XXX found that YYY. However, most of these studies employed simplified models that assume a specific analytic form for the lens and/or connect the shape of the underlying dark matter halo to its parent galaxy. Notable exceptions are XXX. As the quality and quantity of data improve, we can afford relax some of our historical model assumptions and attempt to actually {\it image} the distribution of mass in lensing galaxies and clusters. We refer to this as `non-parametric' mass modelling, where `non-' means that we deliberately employ far more model parameters than data constraints. Searching through our model parameter space, we can then explore model degeneracies and determine which data best break them.

%The majority of lens modelling efforts to date have focussed on `parametric' modelling, where simple theoretically motived models are fit using maximum likelihood methods to the data \citep{2011A&ARv..19...47K, 1993A&A...273..367K, 2010GReGr..42.2151K}. This is useful in that a simple model provides valuable physical insight. However, as data quality improves and the questions we ask of these data become more refined, it is important to re-asses the impact of our model assumptions. `Non-parameteric' methods -- that assume many more parameters than the available data constraints -- are invaluable in this respect. In non-parametric mass modelling, the lens inversion problem is deliberately chosen to be under-constrained \citep{1997MNRAS.292..148S,2005MNRAS.360..477D,2010ApJ...723.1678C,2006MNRAS.367.1209L,2013arXiv1304.2393S}. Instead of finding one `best-fit' model, we must then search for many models consistent with the data, exploring degeneracies between different but plausible\footnotemark\ mass models \citep{2000AJ....120.1654S,2006ApJ...653..936S,2008MNRAS.386..307L}. The key advantages of non-parametric modelling are that we can: (i) explore model degeneracies and marginalise over them; and (ii) we can determine what type and quality of data (including data from stellar population synthesis modelling, stellar kinematics, or other indepedent probes of the mass) are required to measure parameters of interest.
%\footnotetext{The word {\it plausible} encodes the key hidden gremlin in non-parametric methods: constraints extra to the available data (priors) are typically required on the solution space in order to make it finite. However, even rather restrictive priors are far less restrictive than the implicit constraints inherent in parametric mass modelling.}

%that builds on an earlier code \PixeLens\ \citep{1997MNRAS.292..148S,2008ApJ...679...17C}. Like \PixeLens, we

In this paper, we present a new non-parametric strong lens modelling framework \Glass\ that uses an adaptive grid of mass pixels to model the lens, searching through many models consistent with the data to marginalise over model uncertainties. \Glass\ shares some aspects with an earlier code \PixeLens\ \citep{1997MNRAS.292..148S,2008ApJ...679...17C}, but has important new functionality:

@@

We apply \Glass\ to a suite of dynamically realistic mock N-body data to determine what quality of data from strong lensing, stellar mass (derived from populations synthesis modelling) and/or stellar kinematics are required to measure the circularly averaged mass profile of a lens and its shape. We will consider using strong lensing as a cosmological probe, and present the first results from \Glass\ applied to real data, in forthcoming papers.

@@

For sources on-axis (\vec\beta\ = \vec0), the images split into a degenerate Einstein ring with $\theta = \theta_E$. Even in this situation of maximal information (since the source position $\vec\beta$ cannot typically be known), there is a trivial degeneracy between $M(<\theta_E)$ and the angular diameter distances that depend on cosmological parameters (most strongly on the Hubble constant, $D_{L,S} \propto 1/H_0$). For fixed cosmology, a single source splitting into images in the most ideal situation gives us only an enclosed mass. This `steepness' degeneracy (also known as the mass-sheet degeneracy) generalises to the non-circularly symmetric case (CITES). Thus, to measure the mass profile of a lens, we require more information. This can come in the form of another source at different redshift that has its own Einstein radius, giving $M(<\theta_E')$. Or it can come in the form of additional constraints like time delay data between the images, stellar mass estimates from stellar populations synthesis modelling (CITE), stellar kinematics (CITE), or similar. However, even with further information, higher order degeneracies remain. XXX showed that there is a monopole degeneracy that allows additional mass to be deposited in-between images without any optical effect, while YYY find twisting degeneracies exist if ZZZ. These higher order degeneracies have a larger effect on shape measurement than the mass profile but remain largely uncharacterised in the literature. In \S\ref{sec:results}, we will use \Glass\ to determine how well such degeneracies can be broken by data of ever-increasing quality to measure the mass profile and shape of a lens.

@@

>>>>

The majority of lens modelling efforts to date have focussed on `parametric' modelling, where simple theoretically motived models are fit using maximum likelihood methods to the data \citep{2011A&ARv..19...47K, 1993A&A...273..367K, 2010GReGr..42.2151K}. This is useful in that a simple model provides valuable physical insight. However, as data quality improves and the questions we ask of these data become more refined, it is important to re-asses the impact of our model assumptions. `Non-parameteric' methods -- that assume many more parameters than the available data constraints -- are invaluable in this respect. In non-parametric mass modelling, the lens inversion problem is deliberately chosen to be under-constrained \citep{1997MNRAS.292..148S,2005MNRAS.360..477D,2010ApJ...723.1678C,2006MNRAS.367.1209L,2013arXiv1304.2393S}. Instead of finding one `best-fit' model, we must then search for many models consistent with the data, exploring degeneracies between different but plausible\footnotemark\ mass models \citep{2000AJ....120.1654S,2006ApJ...653..936S,2008MNRAS.386..307L}. The key advantages of non-parametric modelling are that we can: (i) explore model degeneracies and marginalise over them; and (ii) we can determine what type and quality of data (including data from stellar population synthesis modelling, stellar kinematics, or other independent probes of the mass) are required to measure parameters of interest.

\footnotetext{The word {\it plausible} encodes the key hidden gremlin in non-parametric methods: constraints extra to the available data (priors) are typically required on the solution space in order to make it finite. However, even rather restrictive priors are far less restrictive than the implicit constraints inherent in parametric mass modelling.}

In this paper, we use a new non-parametric strong lens modelling framework \Glass\ to determine what quality and combination of lensing, stellar mass, and stellar kinematic constraints best constrain the projected mass profile and shape of a gravitational lens. This is particularly timely given the coming explosion in strong lensing and complimentary data. Some significant work has already proceeded in this direction in the literature. \citet{2007ApJ...667..645R} were the first to test the recovery of strong lensing surface density profiles using mock data generated from N-body models {\bf JR. Is this true?}. They found that when including time delays, circularly averaged surface density distributions can sufficiently well-recovered to make interesting comparisons with theoretical models\footnote{Cosmological parameters, however, are more sensitive. \citet{2007ApJ...667..645R} found that over-restrictive model assumptions can lead to significant bias on the Hubble parameter, $H_0$, for example, explaining earlier tensions reported in the literature \citep[e.g.][]{2002astro.ph..4043K}.}. Since then, several groups have used N-body mock data to explore the fidelity of strong lensing reconstruction. \citet{2007MNRAS.380.1729L} explore the recovery of N-body models using a new non-parametric genetic algorithm. Such comparisons were key in the development of their novel `null-space' constraint that includes information about where there a no images in addition to where there are observed ones. XXX did YYY, finding ZZZ {\bf JR. I am sure there are many more references that ought to be added here.}. Here, we expand on these earlier works by considering not only the circularly averaged surface density profile of a lens, but also its shape, higher order measures of the mass recovery (a pixel-by-pixel comparison), and by considering the utility of wrapping in constraints from stellar mass estimates, or stellar kinematics.

We use a new lensing tool, \Glass\, that uses an adaptive grid of mass pixels to model the lens, generating thousands of models to marginalise over uncertainties. \Glass\ shares some heritage with an earlier code, \PixeLens\ \citep{1997MNRAS.292..148S,2008ApJ...679...17C}, but differs in several important respects:

@@

This new functionality allows us to use model strong lens systems while wrapping in additional data constraints. Linear constraints -- like the stellar mass -- are applied simultaneously with the strong lensing constraints; non-linear constraints are applied in a post-processing step, where models are rejected using a likelihood analysis. This latter was prohibitively expensive with \PixeLens\ as only up to $\sim$1000 models could be generated in a reasonable time. With \Glass, we can routinely generate hundreds of thousands of models such that we can afford to discard some if they are inconsistent with independent data. In this first application of \Glass, we apply it to a suite of dynamically realistic mock N-body data to test the recovery of the circularly averaged mass profile of a lens and its shape. We will consider using strong lensing as a cosmological probe, and present the first results from \Glass\ applied to real data, in forthcoming papers.

@@

For sources on-axis (\vec\beta\ = \vec0), the images split into a degenerate Einstein ring with $\theta = \theta_E$. Even in this situation of maximal information (since the source position $\vec\beta$ cannot typically be known), there is a trivial degeneracy between $M(<\theta_E)$ and the angular diameter distances that depend on cosmological parameters (most strongly on the Hubble constant, $D_{L,S} \propto 1/H_0$). For fixed cosmology, a single source splitting into images in the most ideal situation gives us only an enclosed mass. This `steepness' degeneracy (also known as the mass-sheet degeneracy) generalises to the non-circularly symmetric case (CITES). Thus, to measure the mass profile of a lens, we require more information. This can come in the form of another source at different redshift that has its own Einstein radius, giving $M(<\theta_E')$. Or it can come in the form of additional constraints like time delay data between the images, stellar mass estimates from stellar populations synthesis modelling (CITE), stellar kinematics (CITE), or similar. However, even with further information, higher order degeneracies remain. XXX showed that there is a monopole degeneracy that allows additional mass to be deposited in-between images without any optical effect, while YYY find twisting degeneracies exist if ZZZ. These higher order degeneracies have a larger effect on shape measurement than the mass profile but remain largely uncharacterised in the literature. Further degeneracies enter also if there is a strong {\it external} field acting on the lens -- for example from a larger group or cluster environment within which the lens sits (CITES); this {\it shear term} can, in principle, however, be constrained by data (CITES). In \S\ref{sec:results}, we will use \Glass\ to determine how well such degeneracies can be broken by data of ever-increasing quality to measure the mass profile and shape of a lens. We assume throughout that the shear is negligible.

@@

\subsection{Removing models with extra images}

\subsection{A post-processing module for stellar kinematics}
